In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-07 paper/main.tex
More shuffling
revision from Marius I

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular enrichment and clearance. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

@@

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of molecular enrichment and clearance of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of model representation of the intrathecal tracer injection; 
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}(\Cref{fig:results1}C), we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}D--E). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}F). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}G).

@@

The enrichment patterns observed clinically after intrathecal
injection of contrast differ substantially between subjects and
between pathological conditions~\cite{ringstad2018brain,
  eide2021direction, eide2021impaired, eide2022altered}. These
neurological conditions are also associated with alterations in the
pulsatile flow of CSF in the ventricular system and
SAS~\cite{eide2021direction}. Clearly, key physiological factors such
as cardiac pulsatility and respiration easily differ between
individuals and between both pathological and physiological states. We
therefore next asked to what extent -- and how -- variations in CSF
pulsatility would affect the intracranial enrichment characteristics.

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). To examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Methods). For the cardiac contribution, we compute
CSF pressure and flow fields at peak systolic blood inflow,
corresponding to a reduction in CSF space volume at a total rate of
$6\,$ml/s~\cite{baledent2014imaging, causemann2022human} in the SAS
and $0.31\,$ml/s across the lateral ventricle
surface\cite{vinje2019respiratory}. This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

Now, to examine how CSF pulsatility affects the enrichment patterns,
we consider three variations of the baseline: (i) no CSF production,
(ii) reduced pulsatility and thus decreased dispersion ($0.1 \times
R$), and (iii) higher pulsatility with increased dispersion ($10
\times R$). Without CSF production, transport is considerably delayed
(\Cref{fig:csf}N, O). Even after $12$ hours, tracer remains in the
subtentorial regions around the cerebellum, the brain stem and in the
surrounding CSF. Interestingly, the lack of CSF production instead
allows tracer to travel upwards through the ventricular system,
reaching the third ventricle after around $12$ hours. On the other
hand, if the CSF pulsatility is reduced, we observe rapid transport
towards the upper convexity of the cranium, as in the baseline model,
but the tracer spreads exclusively within the anterior regions
(\Cref{fig:csf}P). This feature can be attributed to the CSF flow
bifurcation posterior to the ambient cistern (\Cref{fig:csf}D):
without sufficient diffusion, the tracer is unable to cross into the
posterior SAS. Indeed, with higher dispersion, the tracer moves
through the quadrigeminal cistern and further upwards into the
longitudinal fissure, with also enrichment of the cerebellum
(\Cref{fig:csf}Q).

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--B, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

@@

Summarizing and quantifying these observations
(\Cref{fig:compartmentalization}F), we find that neither the time of
first arrival nor the time-to-peak are substantially affected by the
PVS-CSF permeability at any of the periarterial segments considered
(BA, MCAs, ACAs). However, the concentration difference between the
PVS and its surrounding CSF quickly decreases with higher
permeability, as does the time lag between the concentration peaks in
these domains. Thus, fast transport along the PVS is not contingent
upon a structural compartmentalization of the PVS, whereas a sharp
concentration gradient between PVS and CSF is unlikely without a
restricting barrier.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure4.png}
    \caption{\textbf{Does early perivascular enrichment rely on a structural compartmentalization of the PVS?}
    A) The PVS-CSF membrane permeability regulates the exchange of solutes between the PVS and surrounding CSF spaces; 
    B) The arterial network with highlighted regions of interest; 
    C) 2D slices showing zoom-in to the region surrounding the basal artery after 1, 2, 3, and 4 hours for a low (above) and high (below) PVS-CSF permeability model. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));
    D) As in (C) for regions surrounding (from upper to lower) the ACA-A3, ACA-A2, MCA-R and MCA-L segments;}
    \label{fig:compartmentalization}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Comparison of mean tracer concentration in the PVS (dashed) and over the
outer PVS surface (dotted) for the BA, MCA-R, MCA-L, ACA-2, and ACA3 segments over time for the high (purple) and low permeability (yellow) models;
F) Effect of varying the PVS-CSF permeability on first-time-of-arrival (FTA), time difference ($\Delta t$) between concentration peaks in the PVS and surrounding tissue, time of PVS peak concentration, PVS peak concentration and difference in peak concentration between PVS and surrounding tissue.}
\end{figure}

@@

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement in the SAS reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS in the SAS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

@@

      Total molecular volume & $m_{\rm tot}$ & $0.5$ & mmol & \cite{eide2024functional} \\

>>>>

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular transport. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

@@

\subsection*{In-silico predictions of intracranial transport after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{In-silico modelling of transport of the PVS, SAS and parenchyma.}
     A) Illustration of the model geometry including the CSF-filled spaces (ventricles and SAS) $\Omega_{\rm CSF}$, the parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and veins $\Lambda_{\rm vein}$, as well as their interfaces and boundaries; 
     B) 3D rendering of the computational geometry: the parenchyma, CSF (clipped, blue), arterial network (red), and venous network (blue); 
     C) illustration of model representation of the intrathecal tracer injection; 
     D) in-silico predictions of tracer concentration after 1, 6, 12, and 24 hours (opacity increasing linearly from 0 to 2 mmol/l); 
     E) sagittal, coronal and axial view of in-silico tracer concentrations overlayed on T1-weighted MR image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h; 
     G) total amount of tracer in each compartment over the first 24\,h.}
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
ringstad2018brain, eide2024functional}(\Cref{fig:results1}C), we then predict the spreading of $0.5$ mmol intrathecally injected Gadobutrol after its appearance at the craniocervical junction, represented by tracer inflow across the spinal SAS boundary over the first two hours (\Cref{fig:results1}D--E). 
After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}F). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}G).

@@

The tracer enrichment patterns observed clinically after intrathecal
injection of contrast differ substantially between subjects and
between pathological conditions~\cite{ringstad2018brain,
  eide2021direction, eide2021impaired, eide2022altered}. These
neurological conditions are also associated with alterations in the
pulsatile flow of CSF in the ventricular system and
SAS~\cite{eide2021direction}. Clearly, key physiological factors such
as cardiac pulsatility and respiration easily differ between
individuals and between both pathological and physiological states. We
therefore next asked to what extent -- and how -- variations in CSF
pulsatility would affect the intracranial enrichment characteristics.

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Beginning with the advective, net flow flield driven by 
CSF production, we impose a constant CSF inflow of $400\,$ml/day across the surface
of the lateral ventricles. Simultaneously allowing for efflux across the upper 
convexity yields a total CSF pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). Next, to examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Supplementary information S1.3 for a complete description). 
To estimate the cardiac contribution, we compute the steady state
CSF pressure and flow fields at peak systolic blood inflow.
Modelling the reduction in CSF space volume corresponding to peak systolic
conditions, we assume a total inflow rate of $6\,$ml/s~\cite{baledent2014imaging, causemann2022human}
in the SAS across its outer surface and $0.31\,$ml/s 
across the lateral ventricle surface\cite{vinje2019respiratory}. 
This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

Now, to examine how CSF pulsatility affects the tracer enrichment patterns,
we consider three variations of the baseline: (i) no CSF production,
(ii) reduced pulsatility and thus decreased dispersion ($0.1 \times
R$), and (iii) higher pulsatility with increased dispersion ($10
\times R$). Without CSF production, transport is considerably delayed
(\Cref{fig:csf}N, O). Even after $12$ hours, tracer remains in the
subtentorial regions around the cerebellum, the brain stem and in the
surrounding CSF. Interestingly, the lack of CSF production instead
allows tracer to travel upwards through the ventricular system,
reaching the third ventricle after around $12$ hours. On the other
hand, if the CSF pulsatility is reduced, we observe rapid transport
towards the upper convexity of the cranium, as in the baseline model,
but the tracer spreads exclusively within the anterior regions
(\Cref{fig:csf}P). This feature can be attributed to the CSF flow
bifurcation posterior to the ambient cistern (\Cref{fig:csf}D):
without sufficient diffusion, the tracer is unable to cross into the
posterior SAS. Indeed, with higher dispersion, the tracer moves
through the quadrigeminal cistern and further upwards into the
longitudinal fissure, with also enrichment of the cerebellum
(\Cref{fig:csf}Q).

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--B, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Due to imaging limitiations, the network primarily covers the supratentorial regions, with the lowest network point located approximately $4\,$cm above the craniocervical junction.
Imposing the pressure field induced by CSF production at the periarterial network ends induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

@@

Summarizing and quantifying these observations
(\Cref{fig:compartmentalization}F), we find that neither the time of
first arrival nor the time-to-peak are substantially affected by the
PVS-CSF permeability at any of the periarterial segments considered
(BA, MCAs, ACAs). However, the concentration difference between the
PVS and its surrounding CSF quickly decreases with higher
permeability, as does the time lag between the concentration peaks in
these domains. Thus, fast transport along the PVS is not contingent
upon a structural compartmentalization of the PVS, whereas a sharp
concentration gradient between PVS and CSF is unlikely without a
restricting barrier.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure4.png}
    \caption{\textbf{Does early perivascular enrichment rely on a structural compartmentalization of the PVS?}
    A) The PVS-CSF membrane permeability regulates the exchange of solutes between the PVS and surrounding CSF spaces; 
    B) The arterial network with highlighted regions of interest; 
    C) 2D slices showing zoom-in to the region surrounding the basal artery after 1, 2, 3, and 4 hours for a low (above) and high (below) PVS-CSF permeability model. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF). The cutting planes are normal to the blood vessels;
    D) As in (C) for regions surrounding (from upper to lower) the ACA-A3, ACA-A2, MCA-R and MCA-L segments;}
    \label{fig:compartmentalization}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Comparison of mean tracer concentration in the PVS (dashed) and over the
outer PVS surface (dotted) for the BA, MCA-R, MCA-L, ACA-2, and ACA3 segments over time for the high (purple) and low permeability (yellow) models;
F) Effect of varying the PVS-CSF permeability on first-time-of-arrival (FTA), time difference ($\Delta t$) between concentration peaks in the PVS and surrounding tissue, time of PVS peak concentration, PVS peak concentration and difference in peak concentration between PVS and surrounding tissue.}
\end{figure}

@@

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement in the SAS reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS in the SAS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF).The cutting plane is normal to the blood vessel; 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

For an extended description of the mathematical model and simulation
scenarios, see the Supplementary information S1.

@@

      Total amount of injected tracer & $m_{\rm tot}$ & $0.5$ & mmol & \cite{eide2024functional} \\
